# Introduction {#intro}

## Elements of statistical modelling


Probability and statistics can be characterised as the study of variability.
In particular, statistical inference is the  science of analysing
statistical data, viewed as the outcome of some random process, in order to
draw conclusions about that random process.

Statistical models help us to *understand* the random process by which observed data
have been generated. This may be of interest in itself, but also allows us to make
*predictions* and perhaps most importantly *decisions* contingent on
our inferences concerning the process.

It is also important, as part of the modelling process, to acknowledge
that our conclusions are only based on a (potentially small) sample
of possible observations of the process and are therefore subject to error.
The science of statistical inference therefore involves assessment of the uncertainties
associated with the conclusions we draw.

Probability theory is the mathematics associated with randomness and
uncertainty. We usually try to describe random processes using probability
models. Then, statistical inference may involve estimating any unspecified
features of a model, comparing competing models, and assessing the
appropriateness of a model; all in the light of observed data.

In order to identify 'good' statistical models, we require some principles on which
to base our modelling procedures. In general, we have three requirements of a statistical model

* Plausibility
* Parsimony
* Goodness of fit

The first of these is not a statistical consideration, and
a subject-matter expert usually needs to be
consulted about this. For some objectives, like prediction, 
it might be considered unimportant. Parsimony and goodness of 
fit are statistical issues. Indeed,
there is usually a trade-off between the two and our statistical 
modelling strategies will take
account of this.


## Regression Models
Many statistical models, and all the ones we shall deal 
with in MATH3091, can 
be formulated as *regression* models.

In practical applications, we often distinguish between a *response*
variable and a group of *explanatory* variables. The aim is to determine
the pattern of dependence of the response variable on the explanatory
variables.  A regression model has the general form

> response = function(structure and randomness)

The structural part of the model describes how the response depends 
on the explanatory variables and the random part defines the probability
distribution of the response. Together, they produce the response and the
statistical modeller's task is to 'separate' these out. 

## Example data to be analysed

### `nitric`: Nitric acid

This data set relates to 21 successive days of operation of a plant
oxidising ammonia to nitric acid. The response `yield` is ten times the percentage of ingoing
ammonia that is lost as unabsorbed nitric acid (an indirect measure of the
yield). The aim here is to study how the yield depends on flow of air to the plant (`flow`),
temperature of the cooling water
entering the absorption tower (`temp`) and concentration of nitric acid in the
absorbing liquid (`conc`). These data will be analysed  in worksheet 2 using multiple linear 
regression models.   

```{r nitric, include = FALSE, fig.pos = "tb", fig.cap = "A pairs plot of the `nitric` data"}
nitric <- read.csv("nitric.csv")
pairs(nitric)
```



### `birth`: Weight of newborn babies

This data set contains weights of 24 newborn babies. There are two explanatory 
variables, sex (`Sex`) and gestational age in weeks (`Age`) together with the response
variable, birth weight in grams (`Weight`).  The aim here is to study how birth weight
depends on sex and  gestational age. This data set will be analysed  in
worksheet 3 by using multiple linear regression models including
both categorical and continuous explanatory variables.  


```{r birth, include = FALSE, fig.pos = "tb", fig.cap = "The `birth` data"}
birth <- read.csv("birth.csv")
plot(birth$Age, birth$Weight, xlab="Age (weeks)", ylab="Birthweight (g)",
     col = birth$Sex, pch = birth$Sex)
legend("bottomright", legend = c("Male", "Female"), pch = 1:2, col = 1:2)
```


### `survival`: Time to death

This data set, analysed in worksheet 4, contains survival 
times in 10 hour units (`time`) of 48 rats each allocated to one of 12 combinations
of  3 poisons (`poison`) and 4 treatments (`treatment`). 
The aim here is to study how survival time depends
on the poison and the treatment, and to determine whether there is an interaction
between these two categorical variables.

```{r, include = FALSE, fig.pos = "tb", fig.cap = "The `survival` data"}
survival <- read.csv("survival.csv")
survival$poison <- as.factor(survival$poison)
survival$treatment <- as.factor(survival$treatment)
par(mfrow = c(1, 2))
plot(time ~ poison + treatment, data = survival)
```

### `beetle`: Mortality from carbon disulphide

This data set represents the number of beetles exposed (`exposed`) and number 
killed (`killed`) in eight groups exposed to different doses (`dose`)
of a particular insecticide. Interest is focussed on how mortality is 
related to dose. It seems sensible to model the number of beetles 
killed in each group as the binomial random variable with 
probability of death depending on dose. 
This will be discussed in worksheet 5.

```{r, include = FALSE}
beetle <- read.csv("beetle.csv")
beetle
```



### `shuttle`: Challenger disaster

This data set concerns the 23 space shuttle flights
before the Challenger disaster.  The disaster is thought to have 
been caused by the failure of a number of O-rings, 
of which there were six in total. The data 
consist of four variables, the number of
damaged O-rings for each
pre-Challenger flight (`n_damaged`), 
together with the launch temperature in degrees
Fahrenheit (`temp`), the pressure at which the pre-launch test of O-ring leakage was
carried out (`pressure`) and the name of the orbiter (`orbiter`).
The Challenger launch
temperature on 20th January 1986 was 31F.
The aim is to predict the probability of
O-ring damage at the Challenger launch. 
 This will be discussed in worksheet 6.

```{r, include = FALSE}
shuttle <- read.csv("shuttle.csv")
head(shuttle) 
```

 
### `heart`: Treatment for heart attack

This data set represents the results of a clinical trial
to assess the effectiveness of a thrombolytic (clot-busting) treatment
for patients who have suffered an acute myocardial infarction (heart attack).
There are four categorical explanatory variables, representing

- the `site` of infarction: anterior, inferior or other
- the `time` between infarction and treatment: $\le 12$ or $>12$ hours
- whether the patient was already taking Beta-blocker medication prior to the infarction, `blocker`: yes or no
- the `treatment` the patient was given: active or placebo.

For each combination of these categorical variables, 
the dataset gives the total number of patients (`n_patients`),
and the number who survived for
for 35 days (`n_survived`). The aim is to find out how
these categorical variables affect a patient's chance
of survival. These data will be analysed in worksheet 7.

```{r, include = FALSE}
heart <- read.csv("heart.csv")
head(heart)
```

### `accident`: Road traffic accidents

This example concerns the number of road accidents (`number`) and the 
volume of traffic (`volume`), on each of two roads in Cambridge (`road`),
at various times of day (`time`, taking values `morning`, `midday` or `afternoon`).
We should be able to answer questions like:  

1. Is Mill Road more dangerous than Trumpington Road?
1. How does time of day affect the rate of road accident? 

These issues will be considered in worksheet 8. 

```{r, include = FALSE}
accident <- read.csv("accident.csv")
head(accident)
```


### `lymphoma`: Lymphoma patients
The  `lymphoma` data set represents 30 lymphoma 
patients classified by sex (`Sex`), cell type of lymphoma (`Cell`) and response to 
treatment (`Remis`). It is an example of data which may be represented as
a three-way ($2\times 2\times 2$)
contingency table. The aim here is to study the 
complex dependence structures between the three classifying factors. 
This is taken up in worksheet 9. 

```{r, include = FALSE}
lymphoma <- read.csv("lymphoma.csv")
head(lymphoma)
```
