---
title: "MATH3012: Chapter 1, Lecture 5"
date: ""
author: Helen Ogden
header-includes:
- \usepackage{bm}
- \usepackage{booktabs, multirow}
output:
  beamer_presentation
---


```{r, echo = FALSE}
knitr::opts_knit$set(root.dir = '../datasets')
```

## Recap

Last time, we

- defined the observed information as minus the second derivative
of the loglikelihood.
- define the Fisher information as the expected value of the 
observed information.
- claimed that the variance of the score is equal to the Fisher
information.

Let's revisit this result now.

## A link between the score and expected information
The variance-covariance matrix
of the score vector is equal to the
expected information matrix *i.e.*
\[\text{Var}[U(\bm{\theta})]=\mathcal{I}(\bm \theta)\]
or
\[\text{Var}[U(\bm{\theta})]_{ij}= [\mathcal{I}(\bm \theta)]_{ij}, 
\quad i=1,\ldots ,p, \quad j=1,\ldots ,p\]
provided that

1. The variance exists.
1. The sample space for $\bm Y$ does not depend on $\bm \theta$.

## Proof (continuous $\bm y$ -- in discrete case replace $\int$ by $\sum$)}

For each $i = 1,\ldots, p$ and $j = 1, \ldots, p$,
\begin{align*}
\text{Var}[U(\bm{\theta})]_{ij}&= E[U_i(\bm \theta)U_j(\bm \theta)]\cr
&= \int {{\partial}\over{\partial\theta_i}} \ell(\theta)
{{\partial}\over{\partial\theta_j}} \ell(\theta) f_{\bm Y}(\bm y; \bm \theta) d\bm y\cr
&= \int {{\partial}\over{\partial\theta_i}} \log f_{\bm Y}(\bm y; \bm \theta)
{{\partial}\over{\partial\theta_j}} \log f_{\bm Y}(\bm y; \bm \theta) f_{\bm Y}(\bm y; \bm \theta) d\bm y\cr
&= \int {{{{\partial}\over{\partial\theta_i}}f_{\bm Y}(\bm y; \bm \theta)}\over f_{\bm Y}(\bm y; \bm \theta)}
{{{{\partial}\over{\partial\theta_j}}f_{\bm Y}(\bm y; \bm \theta)}\over f_{\bm Y}(\bm y; \bm \theta)} f_{\bm Y}(\bm y; \bm \theta)d\bm y\cr
&= \int \frac{1}{f_{\bm Y}(\bm y; \bm \theta)}{{\partial}\over{\partial\theta_i}}f_{\bm Y}(\bm y; \bm \theta)
 {{\partial}\over{\partial\theta_j}}  f_{\bm Y}(\bm y; \bm \theta)  d\bm y.
\end{align*}

## Proof 
Now
\begin{align*}
[\mathcal{I}(\bm \theta)]_{ij}&=E\left[-{{\partial^2}\over
{\partial\theta_i\partial\theta_j}} \ell(\theta)\right]\cr
&=\int -{{\partial^2}\over
{\partial\theta_i\partial\theta_j}} \log f_{\bm Y}(\bm y; \bm \theta)  f_{\bm Y}(\bm y; \bm \theta) d\bm y\cr
&=\int -{{\partial}\over{\partial\theta_i}}\left[
{{{{\partial}\over{\partial\theta_j}} f_{\bm Y}(\bm y; \bm \theta)}\over  f_{\bm Y}(\bm y; \bm \theta)}\right]
 f_{\bm Y}(\bm y; \bm \theta) d\bm y\cr
&=\int \left[
-{{{{\partial^2}\over{\partial\theta_i\partial\theta_j}} f_{\bm Y}(\bm y; \bm \theta)}\over  f_{\bm Y}(\bm y; \bm \theta)}
+ {{{{\partial}\over{\partial\theta_i}} f_{\bm Y}(\bm y; \bm \theta)
{{\partial}\over{\partial\theta_j}} f_{\bm Y}(\bm y; \bm \theta)}\over  f_{\bm Y}(\bm y; \bm \theta)^2} \right]
 f_{\bm Y}(\bm y; \bm \theta) d\bm y\cr
&= -{{\partial^2}\over{\partial\theta_i\partial\theta_j}}\int  f_{\bm Y}(\bm y; \bm \theta) d\bm y
+ \int \frac{1}{f_{\bm Y}(\bm y; \bm \theta)}{{\partial}\over{\partial\theta_i}} f_{\bm Y}(\bm y; \bm \theta)
{{\partial}\over{\partial\theta_j}} f_{\bm Y}(\bm y; \bm \theta)  d\bm y\cr
&= \text{Var}[U(\bm{\theta})]_{ij}
\end{align*}


## Maximum likelihood estimation

Maximum likelihood estimation is an attractive method of estimation:

- It is intuitively sensible (choose $\bm \theta$ which makes the observed data
most probable)
- It is general: you just need a probability model and some data!
- Computing the the MLE is often fairly simple.
  Even when the simultaneous equations we obtain by
  differentiating the loglikelihood function are impossible to solve directly,
  solution by numerical methods is usually feasible.
\pause

One more reason to love the MLE... \pause it is
the "best possible" estimator for large sample sizes.

## Asymptotic distribution of the MLE

Suppose that $y_1, \ldots, y_n$ are observations of  independent random variables $Y_1, \ldots, Y_n$,
whose joint p.d.f.  $f_{\bm Y}(\bm y;\bm \theta)=\prod_{i=1}^n f_{Y_i}(y_i;\bm \theta)$ is
completely specified except for the values of an unknown parameter vector
$\bm \theta$, and that
$\hat{\bm \theta}$ is the maximum likelihood estimator of $\bm \theta$.


As $n\to\infty$, the distribution of $\hat{\bm \theta}$ tends to a multivariate
normal distribution with mean vector $\bm \theta$ and variance covariance matrix
$\mathcal{I}(\bm \theta)^{-1}$.

Where $p=1$ and $\bm \theta=(\theta)$, the distribution of the MLE $\hat{\theta}$
tends to $N[\theta,1/{\cal I}(\theta)]$.

## Approximate distribution of the MLE

For 'large enough $n$', we can treat the asymptotic distribution of the MLE as
an approximation. The fact that $E(\hat{\bm \theta})\approx\bm \theta$ means that 
the MLE is *approximately unbiased*
for large samples. 

\pause
The variance of $\hat{\bm \theta}$ is approximately $\mathcal{I}(\bm \theta)^{-1}$. 
It is possible to show that this is the smallest possible variance of any unbiased estimator of 
$\bm \theta$ (this result is called the Cramér--Rao lower bound, which we do not prove here).
Therefore the MLE is the 'best possible' estimator in large samples.


## Sketch proof (one parameter case)

Suppose that $y_1, \ldots, y_n$ are observations of i.i.d. random variables 
$Y_1, \ldots, Y_n$,
whose joint p.d.f. $f_{\bm Y}(\bm y;\theta)=\prod_{i=1}^n f_{Y}(y_i;\theta)$ is
completely specified except for the value of an unknown parameter 
$\theta$, and that
$\hat{\theta}$ is the maximum likelihood estimator of $\theta$.
\pause

We can write the score as
\[u(\theta)={{\partial}\over{\partial\theta}}
\ell(\theta)=\sum_{i=1}^n {{\partial}\over{\partial\theta}} \log f_{Y}(y_i;\theta)\]
so $U(\theta)$ can be expressed as the sum
of $n$ i.i.d. random variables.

\pause
Asymptotically, as $n\to\infty$, by the central limit theorem,
$U(\theta)$ is normally distributed. \pause

But $E[U(\theta)]=0$ and
$\text{Var}[U(\theta)]={\cal I}(\theta)$, so asymptotically 
\[U(\theta) \sim N[0,{\cal I}(\theta)].\]

## Sketch proof (continued)
A Taylor series expansion of $U(\hat{\theta})$ around the true $\theta$
gives
\[U(\hat{\theta})=U(\theta)+(\hat{\theta}-\theta)U'(\theta) +\ldots\]
\pause
Now, $U(\hat{\theta})=0$, and if we approximate $H(\theta)$ by
$E[H(\theta)]=-{\cal I}(\theta)$, and also ignore higher order terms
\[\hat{\theta}=\theta+{1\over{{\cal I}(\theta)}}U(\theta).\]
\pause
As $U(\theta)$ is asymptotically
$N[0,{\cal I}(\theta)]$, $\hat{\theta}$ is asymptotically
$N[\theta,{\cal I}(\theta)^{-1}]$.

## Example (Bernoulli)
If $Y_1, \ldots, Y_n$ are i.i.d. Bernoulli$(p)$ random
variables then asymptotically $\hat p=\bar Y$ has a $N(p,{p(1-p)}/ n)$
distribution.

## Distribution of $\hat p$ ($n = 10$)
```{r, echo = FALSE, fig.height = 4, fig.width = 5}
n <- 10
p <- 0.45
y_samples <- replicate(10000, rbinom(n, 1, p))
y_bar_samples <- apply(y_samples, 2, mean)
plot(ecdf(y_bar_samples), verticals = TRUE, do.points = FALSE, main = "",
     xlab = "p", ylab = "CDF of MLE")
```

## Distribution of $\hat p$ ($n = 10$, $p = 0.45$)
```{r, echo = FALSE, fig.height = 4, fig.width = 5}
n <- 10
p <- 0.45
y_samples <- replicate(10000, rbinom(n, 1, p))
y_bar_samples <- apply(y_samples, 2, mean)
plot(ecdf(y_bar_samples), verticals = TRUE, do.points = FALSE, main = "",
     xlab = "p", ylab = "CDF of MLE")
curve(pnorm(x, mean = p, sd = sqrt(p * (1 - p) / n)), add = TRUE,
      lty = 2, col = 2)
```

## Distribution of $\hat p$ ($n = 100$, $p = 0.45$)
```{r, echo = FALSE, fig.height = 4, fig.width = 5}
n <- 100
p <- 0.45
y_samples <- replicate(10000, rbinom(n, 1, p))
y_bar_samples <- apply(y_samples, 2, mean)
plot(ecdf(y_bar_samples), verticals = TRUE, do.points = FALSE, main = "",
     xlab = "p", ylab = "CDF of MLE")
curve(pnorm(x, mean = p, sd = sqrt(p * (1 - p) / n)), add = TRUE, 
      lty = 2, col = 2)
```

## Distribution of $\hat p$ ($n = 1000$, $p = 0.45$)
```{r, echo = FALSE, fig.height = 4, fig.width = 5}
n <- 1000
p <- 0.45
y_samples <- replicate(10000, rbinom(n, 1, p))
y_bar_samples <- apply(y_samples, 2, mean)
plot(ecdf(y_bar_samples), verticals = TRUE, do.points = FALSE, main = "",
     xlab = "p", ylab = "CDF of MLE")
curve(pnorm(x, mean = p, sd = sqrt(p * (1 - p) / n)), add = TRUE,
      lty = 2, col = 2)
```

## Conclusion

- We have found one of the most important results about the 
MLE: it has asymptotic 
$N(\theta, [{\cal I}(\theta)]^{-1})$ distribution.
- So for large sample sizes, the MLE is essentially 
the best possible estimator we could construct.
- Next time we'll see how we can use the asymptotic distribution 
to find confidence intervals for the parameters, and to 
test hypotheses.
