---
title: "MATH3012 problem sheet 4"
author: ""
date: ""
output: pdf_document
fontsize: 12pt
header-includes:
- \newcommand{\benum}{\begin{enumerate}}
- \newcommand{\eenum}{\end{enumerate}}
- \usepackage{bm}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = file.path('..', 'datasets'))
```


\benum
\item The time to failure ($Y$) of a certain type of electrical component is
thought to follow an exponential distribution, with probability density
of the form
\[f_Y(y;\lambda) = \exp\left(-\lambda y\right), \quad
y>0;\quad\lambda>0.\]

It is believed that the failure rate of a component $\lambda$ is related
to its electrical resistance ($x$) by the relationship
\[\lambda = \beta_1+\beta_2 x.\]
Suppose that $y_1,...,y_n$  are observations of the
times to failure, $Y_1,...,Y_n$ for $n$ such components with corresponding
resistances $x_1,...,x_n$.

\benum
\item Write down the likelihood in terms of $\beta_1$ and
$\beta_2$ and hence derive a pair of
simultaneous equations, the solutions of which are the maximum likelihood
estimates.

\item Calculate the observed and expected information matrices.
Are the Newton-Raphson and the Fisher scoring methods identical
for this problem? Justify your answer.

\eenum

\item
Suppose $Y_i \sim \text{geometric}(p_i)$, the 
geometric distribution as studied in Question 2 of problem
sheet 3. We want to model how $p_i$ depends on explanatory
variables $\bm x_i$.
\benum
\item Assuming a GLM with canonical link, write down a formula for $p_i$ in terms of $\bm x_i$.
\item Derive an expression for the scaled deviance for this model,
writing $\hat \mu_i$ for the estimate of $\mu_i = E(Y_i)$ under the model.
Write an expression for $\hat \mu_i$ in terms of $\hat{\bm \beta}$, the MLE
of $\beta$ under the model.
\eenum

\item
We return to the `beetle` data studied in computer worksheet 5,
with observations on $n = 8$ groups of beetles.
There we considered the model:
```{r, include = FALSE}
beetle <- read.csv("beetle.csv")
beetle$prop_killed <- beetle$killed / beetle$exposed
```
```{r}
beetle_glm <- glm(prop_killed ~ dose, data = beetle, family = binomial,
                  weights = exposed)
```
We could have also considered a model with quadratic dependence on `dose`:
```{r}
beetle_glm_quad <- glm(prop_killed ~ dose + I(dose^2), data = beetle, 
                       family = binomial, weights = exposed)
```
\benum
\item Write down mathematical expressions for the two models. Show that 
`beetle_glm` is nested with `beetle_glm_quad`, and write down the null hypothesis
$H_0$ and the alternative hypothesis $H_1$ you would use for comparing the models.
\item Consider the following output of a `summary` call:
```{r}
summary(beetle_glm)
```
What is the scaled deviance for `beetle_glm`? 
\item The scaled deviance for `beetle_glm_quad` is $3.1949$. 
Calculate the log likelihood ratio test statistic $L_{01}$ 
for testing $H_0$ against $H_1$.
Under $H_0$, what is the distribution of this statistic? Hence 
conduct a hypothesis test of $H_0$ against $H_1$, and make a conclusion
about which model you prefer. 
\eenum

\eenum