---
title: 'MATH3012 worksheet 8'
author: ""
date: ""
output: pdf_document
fontsize: 12pt
---

```{r setup, include=FALSE, purl = FALSE}
knitr::opts_chunk$set(eval = FALSE)
knitr::opts_knit$set(root.dir = file.path('..', 'datasets'))
```

## Analysis of `accident` data


The data in `accident.csv` concerns the number of road accidents (`number`) and the 
volume of traffic (`volume`), on each of two roads in Cambridge (`road`),
at various times of day (`time`, taking values `morning`, `midday` or `afternoon`).
We would like to answer questions like:

1. Is Mill Road more dangerous than Trumpington Road?
1. How does time of day affect the rate of road accidents? 


Let $Y_{ij}$ denote the number of accidents and $\mu_{ij}$ denote the expected
number of accidents for the $i$th road at the $j$th time.
We assume
$$
Y_{ij} \sim \text{Poisson}(\mu_{ij}), \ i=1, 2 \text{ for Road},
j=1, 2, 3 \text{ for time}.
$$


We might reasonably expect the number of accidents to depend
on traffic `volume`, $v_{ij}$.  It is better to work with log volume
rather than volume itself since those are very high.

 Use the Poisson GLM with canonical log link.
$$
\log (\mu_{ij}) = \mu + \alpha_i + \beta_j + \gamma \log v_{ij}.
$$
This is equivalent to assuming:
$$
\mu_{ij} = \text{constant} \times \text{$i$th road effect}
  \times \text{$j$th time effect}  \times \text{volume}^{\gamma}
$$
Read the data from the file `accident.csv` into a data frame called 
`accident`.
```{r, include = FALSE}
# Analysis of accident data

# read in the data
accident <- read.csv("accident.csv")
```

By default, `R` will set $\alpha_1=0$ and $\beta_1=0$.   Thus
$\alpha_2=0$ if the roads are equally risky,
$\beta_2$ represents difference between
time 2 and 1,  and $\beta_3$ represents difference between
time 3 and 1.

We can fit this model in `R` with:
```{r, include = FALSE}

# fit a Poisson log-linear model to the data:
```

```{r}
acc_glm <- glm(number ~ road + time + log(volume), family = poisson,
               data = accident)
summary(acc_glm)
```
What do you conclude about how the accident rate depends on 
the road, the time of day and the volume of traffic?

```{r, include = FALSE}
# The output seems to say Mill road is more dangerous than
# Trumpington road. The mornings and afternoons are about as
# dangerous as each other and each is quite a lot more dangerous than
# the middle of the day. The accident rate has a strong dependence on
# the traffic volume.
```
Does the scaled deviance indicate an problem with 
the fit of the model?

```{r, include = FALSE}

# From the summary output, the scaled deviance is 1.88.
# acc_glm has 5 free parameters, and there are n = 6
# data points, so if the model is correct, we expect
# the scaled deviance to have chi^2_1 distribution.
# The 95% point of a chi^2_1 is
qchisq(0.95, df = 1)
# or 3.84. Since the scaled deviance is less than this,
# there is no evidence of poor model fit from this measure.

```

Conduct hypothesis tests to check whether you could drop any
terms from the model.

```{r, include = FALSE}
# Consider model without volume:
acc_glm_rt <- glm(number ~ road + time, family  = poisson,
                  data = accident)
anova(acc_glm_rt, acc_glm, test="LRT")
# prefer the model with volume

# Consider model without time:
acc_glm_rv <- glm(number ~ road + log(volume), family = poisson,
                  data = accident)
anova(acc_glm_rv, acc_glm, test="LRT")
# It is borderline whether or not to include time

# Consider model without road:
acc_glm_tv <- glm(number ~ time + log(volume), family = poisson,
                  data = accident)
anova(acc_glm_tv, acc_glm, test="LRT")
# prefer the model with road.

```

## Analysis of Hodgkins data

The data in `hodgkins.csv` concerns 538 patients with Hodgkin's disease, who have been
cross-classified according to two factors, `type` the histological type of
their disease (4 levels) and `rtreat`, their response to treatment
(3 levels).

Read the data from the file `hodgkins.csv` into a data frame called 
`hodgkins`.
```{r, include = FALSE}
# Analysis of Hodgkins data

# read in the data:
hodgkins <- read.csv("hodgkins.csv")
```

We could display this data in a contingency table using
```{r, include = FALSE}

# Display the data in contingency table form:
```
```{r}
xtabs(count ~ type + rtreat, data = hodgkins)
```

Classification data are extremely common,
and can be modelled very effectively
using generalised linear models.
The observations of the response variable are taken
to be the counts (in this case the 12 patient totals) and a generalised
linear model is used to determine how the expected counts
depend on any explanatory
variables (in this case, the factors `type` and `rtreat`).
The dataset as it is presented in the original `hodgkins`
data frame is already in the correct format to fit
such a generalised linear model.

Counts are non-negative integers, so one approach is to treat
them as observations of Poisson random variables.
The canonical link function is then the $\log$ function, and Poisson generalised
linear models with the log link are called *log-linear models*.
A possible log-linear model for this data set is:
\[Y_i\sim\text{Poisson}(\mu_i),\quad
  \log\mu_i=\alpha+\beta_T(t_i)+\beta_R(r_i)\qquad i=1,\ldots ,12,\]
where $Y_i$ is the $i$th count and $t_i$
and $r_i$ are the corresponding levels of `type` and `rtreat`.
We can fit this model in `R` with:
```{r, include = FALSE}

# Fit a log-linear model with main effects only:
```
```{r}
hod_main <- glm(count ~ type + rtreat, family = poisson, data = hodgkins)
```

This model only contains the "main effects"
`type` and `rtreat`. 
Alternatively, we could fit a model with an interaction
between these two main effects, which is the saturated model:
```{r, include = FALSE}

# Fit a log-linear model with interaction between main effects:
```
```{r}

hod_int <- glm(count ~ type * rtreat, family = poisson, data = hodgkins)
```
You can find the estimated cell counts
under the models with:
```{r, include= FALSE}

# Find estimated cell counts under the two models:
```{r}
fitted(hod_main)
fitted(hod_int)
```
What do you notice about the estimated cell counts under the saturated
model `hod_int`, compared with the observed counts?
```{r, include = FALSE}
# For the saturated model, the estimated cell counts are equal
# to the observed cell counts
```

Which model do you prefer?
```{r, include = FALSE}

# Compare the two models:
anova(hod_main, hod_int, test = "LRT")

# There is strong evidence to prefer hod_int, and we
# conclude that response to treatment is dependent on
# the type of the disease.
```
By comparing these two models, 
we are determining whether `type` and `rtreat` are independent,
or whether there is significant evidence of association.
If we need the interaction term,
then this provides evidence that the 
cross-classifying factors are not independent of one another.

Is there any evidence of dependence between
response to treatment and type of disease?
