---
title: 'MATH3012 worksheet 6'
author: ""
date: ""
output: pdf_document
fontsize: 12pt
header-includes:
- \newcommand{\benum}{\begin{enumerate}}
- \newcommand{\eenum}{\end{enumerate}}
---

```{r setup, include = FALSE, purl = FALSE}
knitr::opts_chunk$set(eval = FALSE)
knitr::opts_knit$set(root.dir = file.path('..', 'datasets'))
```


Read the data from the file `shuttle.csv` into a data frame called 
`shuttle`.
```{r, include = FALSE}
# read in the data
shuttle <- read.csv("shuttle.csv")
```


This data set concerns the 23 space shuttle flights
before the Challenger disaster.  The disaster is thought to have 
been caused by the failure of a number of O-rings, 
of which there were six in total. The data 
consist of four variables, the number of
damaged O-rings for each
pre-Challenger flight (`n_damaged`), 
together with the launch temperature in degrees
Fahrenheit (`temp`), the pressure at which the pre-launch test of O-ring leakage was
carried out (`pressure`) and the name of the orbiter (`orbiter`).

The Challenger launch
temperature on 20th January 1986 was 31F.
By fitting generalised linear models to this data, we will predict the probability of
O-ring damage at the Challenger launch. 

We need to add a column `n` to the data for the total number of O-rings on each shuttle,
which should be $6$ in all cases:
```{r, include = FALSE}

# add a column to the data for the total number of O-rings on each shuttle:
```
```{r}
shuttle$n <- rep(6, nrow(shuttle))
```

Make three plots, of the 
proportion of O-rings which were damaged against
`temp`, `pressure` and `orbiter`.
```{r, include = FALSE}
# Make plots of proportion of O-rings damaged against
# temp, pressure and orbiter:
plot(n_damaged / n ~ temp + pressure + orbiter, data = shuttle)
```
Here `orbiter` is a factor, and `temp`  and `pressure` are continuous
covariates, and your plots should be appropriate to the type
of explanatory variable.
Which of these explanatory variables do you think will
be important for predicting the proportion of damaged
O-rings?

We first find the best model. We can fit a logistic regression
model for the proportion of damaged O-rings against
`temp`, `pressure` and `orbiter` with
```{r, include = FALSE}

# Fit a logistic regression of proportion of damaged O-rings
# against temp, pressure and orbiter:
```
```{r}
shuttle_glm1 <- glm(n_damaged / n ~ temp + pressure + orbiter,
                    data = shuttle, family = binomial, weights = n)
```
We can check the `summary` to find estimates of the parameters,
and to conduct hypothesis tests about whether each individual parameter
is needed in the model:
```{r}
summary(shuttle_glm1)
```
We can also use `anova` with the `test = "LRT"` argument 
to conduct a log-likelihood ratio test of whether each 
explanatory variable is needed in the model:
```{r, include = FALSE}

# conduct likelihood ratio tests for the presence of each explanatory variable:
```
```{r}
anova(shuttle_glm1, test = "LRT")
```

What do you conclude about which 
explanatory variables you should keep in the model?
```{r, include = FALSE}
# So we should keep temp in the model, but we do not need other variables.
```

We could fit a logistic regression on `temp` alone with
```{r, include = FALSE}
# Fit a logistic regression on temp alone:
```
```{r}
shuttle_glm2 <- glm(n_damaged / n ~ temp, data = shuttle, family = binomial, 
                    weights = n)
```
```{r, include = FALSE}
summary(shuttle_glm2)
```
We might also consider whether there might be a higher-order polynomial 
dependence of the
linear predictor on `temp`, rather than the linear dependence
in `shuttle_glm2`. For example, we could fit a model
with a linear predictor
which is quadratic in `temp`:
```{r, include = FALSE}

# Fit a logistic regression with an linear predictor which is quadratic
# in temp:
```
```{r}
shuttle_glm3 <- glm(n_damaged / n ~ temp + I(temp^2),
                    data = shuttle, family = binomial, weights = n)
```
Do you prefer `shuttle_glm2` or `shuttle_glm3`?
```{r, include = FALSE}
summary(shuttle_glm3)
# The coefficient for I(temp^2) is not required in the model,
# so prefer shuttle_glm2.
```

We would like to use our preferred model to predict the probability of
O-ring damage at 31F.
The command
```{r, include = FALSE}

# predict the probability of O-ring failure at 31F:
```
```{r}
newdata <- data.frame(temp = 31)
pred_glm2 <- predict(shuttle_glm2, newdata = newdata, 
                     type = "response", se.fit = TRUE)
pred_glm2
```
```{r, include = FALSE}
# The predicted probability of O-ring failure at 31F is 0.82
```
We have the predictions and their standard errors, so we can obtain an asymptotic 95% confidence interval:
```{r, include = FALSE}

# find a asymptotic 95% confindence interval for the probability of 
# O-ring failure at 31F:
```
```{r}
CI <- c(pred_glm2$fit - qnorm(0.975) * pred_glm2$se.fit, 
        pred_glm2$fit + qnorm(0.975) * pred_glm2$se.fit)
CI
```

What do you notice about the interval obtained?
```{r, include = FALSE}
# The upper limit of the confidence interval is greater than 1,
# which is not sensible.
```

Try rerunning this code to find a prediction and confidence interval 
without `type = "response"`. 
This does the prediction on the scale of the linear predictor. 
Can you work out how to back-transform the prediction and confidence 
interval to the scale of the data? Is the prediction the same as before?
Is the confidence interval the same as before?

```{r, include = FALSE}

# predict the linear predictor at 31F:
pred_glm2_eta <- predict(shuttle_glm2, newdata = newdata, se.fit = TRUE)

# find a confidnece interval for the linear predictor at 31F:
CI_eta <- c(pred_glm2_eta$fit - qnorm(0.975) * pred_glm2_eta$se.fit, 
            pred_glm2_eta$fit + qnorm(0.975) * pred_glm2_eta$se.fit)
CI_eta

# to find the predicted probability, apply the inverse of the logit link,
# We have g(x) = log(x / 1 - x)
# So g^{-1}(x) = exp(x) / (1 + exp(x)) 
exp(pred_glm2_eta$fit) / (1 + exp(pred_glm2_eta$fit))
# The predicted probability of O-ring failure is 0.82, as before.

exp(CI_eta) / (1 + exp(CI_eta))
# This is different from previous confidence interval.
# The new confidence interval is more sensible than the previous
# one, as the upper limit is less than one.
```

Can you modify the code from worksheet 5 to add a curve
giving the predicted probability of O-ring damage
at each `temp` to your plot
of proportion of damaged O-rings against `temp`? 
Extend the x-axis beyond the range of the original data, to between
30F and 100F
```{r, include = FALSE}

# Plot proportion of O-rings damaged against temp, with a fitted
# probability curve added:
newdata <- data.frame(temp = seq(30, 100, length = 30))
plot(n_damaged / n ~ temp, data = shuttle, ylim = c(0, 1), xlim = c(30, 100))
lines(newdata$temp, predict(shuttle_glm2, newdata = newdata, type = "response"), 
      col = 2)
```

All of these predictions
should be treated wih extreme caution,
as we are extrapolating far outside the range of the observed data.
However, in the presence of such high uncertainty, it may have
been unwise to launch at such a low temperature. 
