---
title: 'MATH3012 worksheet 4: Linear models with factors'
author: ""
date: ""
output: pdf_document
fontsize: 12pt
header-includes:
- \newcommand{\benum}{\begin{enumerate}}
- \newcommand{\eenum}{\end{enumerate}}
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(eval = TRUE)
knitr::opts_knit$set(root.dir = file.path('..', 'datasets'))
```

## Explore the data

Load the data in `survival.csv` into a variable `survival`.
```{r, include = FALSE}
survival <- read.csv("survival.csv")
```

This dataset represent survival times in 10 hour units (`time`) of 48
animals each allocated to one of 12 combinations of 3 poisons (`poison`)
and 4 treatments (`treatment`).
The aim of the experiment is to determine the dependence of survival time
on the explanatory variables poison and treatment.

The values of these explanatory variables are categories.
We refer to categorical explanatory variables as `factors`.
The values assigned to the variables are just labels distinguishing
between categories.
Even if we used numerical labels, their values would have no
intrinsic meaning. Therefore, we cannot include such variables directly
in a linear model. We need to convert the two explanatory variables into
factors.
```{r}
survival$poison <- as.factor(survival$poison)
survival$treatment <- as.factor(survival$treatment)
```

Plot the data. Issue 
```{r}
plot(time ~ poison + treatment, data = survival)
```
You can plot both on the same
graph sheet if you first issue `par(mfrow = c(2,1))` and then the plot command.
Can you order the poison types? How about the treatments?

## Modelling with factors

To model the dependence of the response on an explanatory factor,
we include a model coefficient for each level of the factor.
For example,  a model which allows survival time  to depend on poison type
might be written as
$$
Y_i=\mu+\alpha(p_i)+\epsilon_i,\quad i=1,\ldots ,n\eqno{(1)}
$$
where $Y_i$ is the survival `time` and $p_i$ is the type of `poison` for the $i$th observation.
Here $\alpha(p_i)$ takes three different values depending on the three possible
levels of the factor `poison`.

Note that this model is overparameterised, in that it would be possible
to add some fixed constant to each $\alpha(p_i)$ and subtract the same
constant from $\mu$ and end up with exactly the same dependence of  `time` on `poison`.
Therefore, in practice, we impose a constraint on
$\alpha(p_i)$. The easiest to interpret is to constrain $\alpha(p)$ to be equal to 0
at the first  level of `poison`. 
Which one is the first level of `poison`? Issue the command
```{r}
levels(survival$poison)
```
Thus $\alpha(p_i)=0$ when $p_i=1$.
By default, `R` will set the first level of
all categorical factors to zero in the `lm` function. 


## Models

The model $(1)$ is then fitted by
```{r}
survival_lm1 <- lm(time ~ poison, data = survival)
```
Factors are included in the standard linear model structure by
creating a dummy variable for every level of the factor. The dummy variable
at level $f$ of a factor $F$ takes the value 1 for every observation
where $F=f$ and 0 for all other observations.
The dummy variables are incorporated in the design matrix in the usual way.
The model coefficient $\alpha(f)$ corresponds to the dummy variable
at level $f$.

Factorial models may include more than one factor.
For example, we can model the dependence of `time` on `treatment` and `poison`
using the linear model
$$
Y_i=\mu+\alpha(p_i)+\beta(t_i)+\epsilon_i,\quad i=1,\ldots ,n,
$$
where $t_i$ is the `treatment` for the $i$th observation.
This can be fitted in `R` with
```{r}
survival_lm2 <- lm(time ~ poison + treatment, data = survival)
```

This is an additive model, which assumes that the effect of a treatment is
the same for every poison, and conversely that the
effects of a poison are the same for every treatment.
These effects of poison and treatment are called *main effects*.

## Models with interactions

The additive model may not be appropriate.
Different poisons may respond differently in
combination with different treatments.
We call this *interaction*.
A model with interaction is
$$
Y_i=\mu+\alpha(p_i)+\beta(t_i)+\gamma(p_i,t_i)+\epsilon_i,\quad i=1,\ldots ,n.
$$
We require constraints on the interaction parameters $\gamma$, similar to those
on the main effects.
Usually, we constrain $\gamma(p_i,t_i)$ to be equal to 0 for all combinations
where either `poison` or `treatment` are at their first level.

The model with interaction can be fitted using
```{r}
survival_lm3 <- lm(time ~ poison + treatment + poison:treatment, data = survival)
```

where `poison:treatment` indicates an interaction between `poison` and `treatment`.
Do not attempt to include an interaction term without the
corresponding main effects (the model is meaningless).

Note that in factorial models it rarely makes sense to model some
of the levels of a factor but not others. Therefore, when we add
an explanatory factor $F$ into the model, we are adding $l_F-1$
extra coefficients, where $l_F$ is the number of levels of
the factor concerned.
As usual, we can use an $F$ test to determine whether adding a factor leads
to a significant improvement in fit.

## Further exercises

\benum
\item By comparing models using $F$ tests (remember the `anova`
command) determine which model you prefer. Is it a good model?
\item
Now transform the response to death rate (reciprocal of
survival time). Try modelling death rate using factorial models.
Which model do you prefer now?
\item
In each case, write down your model and the corresponding
parameter estimates. Overall, what is your preferred model.
What conclusions do you draw about the effects of
the different poisons and treatments?
\eenum
