---
title: 'MATH3012 worksheet 7'
author: ""
date: ""
output: pdf_document
fontsize: 12pt
header-includes:
- \newcommand{\benum}{\begin{enumerate}}
- \newcommand{\eenum}{\end{enumerate}}
---

```{r setup, include = FALSE, purl = FALSE}
knitr::opts_chunk$set(eval = FALSE)
knitr::opts_knit$set(root.dir = file.path('..', 'datasets'))
```

Read the data from the file `heart.csv` into a data frame called 
`heart`.
```{r, include = FALSE}
# read in the data
heart <- read.csv("heart.csv")
```

This dataset contains the results of a clinical trial
to assess the effectiveness of a thrombolytic (clot-busting) treatment
for patients who have suffered an acute myocardial infarction (heart attack).

The explanatory variables are:

- the `site` of infarction: anterior, inferior or other. 
- the `time` between infarction and treatment: $\le 12$ or $>12$ hours
- whether the patient was already taking Beta-blocker medication prior to the infarction, `blocker`: yes or no
- the `treatment` the patient was given: active or placebo.

For each combination of these categorical variables, 
the dataset gives the total number of patients (`n_patients`),
and the number who survived for
for 35 days (`n_survived`).

Make plots of the proportion of patients who survived for 
35 days against each of the explanatory variables.
```{r, include = FALSE}

# plot the proportion of patients who survived against 
# each of the explanatory variables
plot(n_survived / n_patients ~ site + time + blocker + treatment,
     data = heart)
```

We abbreviate `site` to $S$, `time` to $T$, `blocker` to $B$ and
`treatment` to $R$ in our mathematical descriptions of various
possible models for this data.

Recall from Worksheet 4 that we refer to categorical explanatory variables
as *factors*, and factorial models can include *main effects*
and *interactions*.
For example, a logistic regression model which allows `n_survived` to
depend on `site`, `treatment` and an interaction between these two factors is
$$
Y_i|n_i, p_i\sim\;{\rm Binomial}(n_i,p_i),\ \ \
\log\left({{p_i}\over{1-p_i}}\right)=\alpha+\beta_S(s_i)
+\beta_{R}(r_i)+\gamma_{S,R}(s_i,r_i), \ \  i=1,\ldots ,24.
$$
where $s_i$ is the level of `site`,
and $r_i$ is the level of `treatment`
for the $i$th observation; $p_i$ is the probability of
surviving at least 35 days and $n_i$ is the number of patients in the $i$th group.
Here $\beta_{S}$, the main effect of `site` takes different values
depending on the level of `site`, so in principle $\beta_{S}$ takes three values, 
$\beta_{S}(\text{anterior})$, $\beta_{S}(\text{inferior})$ and $\beta_{S}(\text{other})$.
Similarly,  $\beta_{R}$ depends on the level of
`treatment` (two values) and $\gamma_{S,R}$ depends jointly on the levels of
`site` and `treatment` (six values).

We could fit this model in `R` with
```{r, include = FALSE}

# Fit a logistic regression of proportion of patients who survive
# on site, treatment, and an interactino between site and treatment:
```
```{r}
mod_SR <- glm(n_survived / n_patients  ~ site + treatment + site:treatment,
                    family = binomial, weights = n_patients, data = heart)
summary(mod_SR)
```

By default, `R` constrains any main effect to be equal to 0 at the
first level of a factor. If the factor levels
are not labelled numerically, `R` interprets the 'first' level
to be the first in alphabetical order. In this case,
the first level of `site` is "anterior", and the first
level of `treatment` is "active", so the intercept in `mod_SR`
represents the linear predictor for patients with an infarction
in the anterior site, who are given the active treatment.

When specifying models with interactions between factors,
one rule which must be followed is
the *principle of marginality* which states that 'whenever an interaction
is present in a model, all *marginal* main effects and interactions
(those which correspond to 'subsets') must also be present.
For example, if we include the `site:treatment` interaction, then the main
effects of `site` and `treatment` must also be in the model, as 
in `mod_SR`.
A useful shorthand permitted by `R`
is `site*treatment` which represents the interaction `site:blocker`
together with all its marginal terms.

```{r, include = FALSE}

# Alternatively, specify the main and interaction effects in one term:
```
```{r}
mod_SR_2 <- glm(n_survived / n_patients  ~ site * treatment,
                family = binomial, weights = n_patients, data = heart)
```
Check that `mod_SR` and `mod_SR_2` are equivalent.
```{r, include = FALSE}

# The same terms are included in either case:
summary(mod_SR)
summary(mod_SR_2)
```


The concept of interaction can be extended when we have three or more
factors. For example, a *three factor interaction* allows the
response to depend jointly on three factors. For example, we could
fit a three factor interaction between `site`, `blocker` and
`treatment` with:
```{r, include = FALSE}

# Fit a three factor interaction between site, blocker and treatment:
```
```{r}
mod_SBR <- glm(n_survived / n_patients  ~ site * blocker * treatment,
               family = binomial, weights = n_patients, data = heart)
```

By default, `R`
constrains any interaction to be equal to 0 for all
combinations where any of the factors are at their first level.
Hence, the main effect `site` involves $3-1 = 2$ free
coefficients, the interaction `site:treatment` involves $(3-1)(2-1) = 2$ free
coefficients, `site:blocker:treatment` involves $(3-1)(2-1)(2-1) = 2$ free
coefficients, and so on.
By the principle of marginality, if we include the three factor interaction `site:blocker:treatment`,
the main effects
`site`, `blocker` and `treatment`, and the interactions `site:blocker`, `site:treatment`
and `blocker:treatment` must all be present. Specifying the model with
`site*blocker*treatment` ensures that all of these terms are included.

There are many possible models that we could consider, with many
possible interaction terms. Let's start with a model including a main effect
for each explanatory variable, but no interaction terms:

```{r, include = FALSE}

# fit a logistic regression with all main effects, but no interactions:
```
```{r}
mod_S_T_B_R <- glm(n_survived / n_patients  ~ site + time + blocker + treatment,
                   family = binomial, weights = n_patients, data = heart)
```
Starting at this model, conduct hypothesis tests to see whether you could
drop any of the main effects.
For the remaining main effects, try adding interactions between pairs of
variables, and conduct hypothesis test of whether or not
each of these possible two factor interactions is needed.
If you include more than one two factor interaction,
check to see if any three factor interactions are required.

```{r, include = FALSE}

# do hypothesis test of each main effect:
anova(mod_S_T_B_R, test = "LRT")
# We should include site, blocker and treatment, but don't need to include time

mod_S_B_R <- glm(n_survived / n_patients  ~ site + blocker + treatment,
                       family = binomial, weights = n_patients, data = heart)
anova(mod_S_B_R, test = "LRT")
# We should keep all of the remaining main effects.

# We should now consider potential interactions between the main effects.
# Possible two factor interactions: site*blocker, site*treatment, blocker*treatment.

mod_SB_R <- glm(n_survived / n_patients  ~ site*blocker + treatment,
                family = binomial, weights = n_patients, data = heart)
anova(mod_S_B_R, mod_SB_R, test = "LRT")
# interaction site:blocker not needed

mod_SR_B <- glm(n_survived / n_patients  ~ site*treatment + blocker,
                family = binomial, weights = n_patients, data = heart)
anova(mod_S_B_R, mod_SR_B, test = "LRT")
# interaction site:treatment not needed

mod_S_BR <- glm(n_survived / n_patients  ~ site + blocker*treatment,
                family = binomial, weights = n_patients, data = heart)
anova(mod_S_B_R, mod_S_BR, test = "LRT")
# interaction blocker:treatment not needed

# so we prefer mod_S_B_R as our final model
```

At the end of this process, using a $5\%$ significance level,
you should find that the preferred model is 
$$
Y_i|n_i, p_i\sim\;{\rm Binomial}(n_i,p_i),\ \ \
\log\left({{p_i}\over{1-p_i}}\right)=\alpha+\beta_S(s_i)+\beta_B(b_i)
+\beta_{R}(r_i), \ \  i=1,\ldots ,24.
$$
where $s_i$ is the level of `site`, $b_i$ is the level of `blocker`,
and $r_i$ is the level of `treatment` for the $i$th observation.

```{r}
mod_final <- glm(n_survived / n_patients  ~ treatment + blocker + site,
                 family = binomial, weights = n_patients, data = heart)
summary(mod_final)
```

The scaled deviance 
(called "Residual deviance" in the output of `summary(mod_final)` above)
provides one check of the model fit.
In this case the scaled deviance 
is 15.86, which is less than
$30.14$, the $95\%$ point of $\chi^2_{19}$, so the scaled deviance
does not provide any evidence of poor model fit.

Consider a patient who had an anterior infarction and was not
previously taking Beta-blockers. What is your estimate of the
probability they will survive for 35 days if they are given
the active treatment? What is your estimate of this probability
if they are given the placebo? Find confidence intervals
for both probabilities.
```{r, include = FALSE}

# Make predictions for a patient with anterior infarction, no beta blockers:
newdata_active <- data.frame(site = "anterior", blocker = "no", treatment = "active")
p_hat_active <- predict(mod_final, newdata = newdata_active, type = "response")
# estimated probability of survival given active treatment is 90%

newdata_placebo <- data.frame(site = "anterior", blocker = "no", treatment = "placebo")
p_hat_placebo <- predict(mod_final, newdata = newdata_placebo, type = "response")
# estimated probability of survival given placebo treatment is 87%

# Find a confidence interval for probability of survival given active treatment:
pred_eta_active <- predict(mod_final, newdata = newdata_active, se.fit = TRUE)
CI_eta_active <- c(pred_eta_active$fit - qnorm(0.975) * pred_eta_active$se.fit, 
                   pred_eta_active$fit + qnorm(0.975) * pred_eta_active$se.fit)
CI_p_active <- exp(CI_eta_active) / (1 + exp(CI_eta_active))

# Find a confidence interval for probability of survival given placebo treatment:
pred_eta_placebo <- predict(mod_final, newdata = newdata_placebo, se.fit = TRUE)
CI_eta_placebo <- c(pred_eta_placebo$fit - qnorm(0.975) * pred_eta_placebo$se.fit, 
                   pred_eta_placebo$fit + qnorm(0.975) * pred_eta_placebo$se.fit)
CI_p_placebo <- exp(CI_eta_placebo) / (1 + exp(CI_eta_placebo))
```

Qualitatively this model suggests that probability of 35 day survival
is enhanced by the thrombotic treatment ($\text{logit}(p)$ and therefore $p$ is
significantly lower for the placebo). As we see from the estimated
probabilities, while this treatment effect is significant, 
the size of the change in probability of survival is fairly small.
We also see that probability of survival is significantly higher for those whose
site of infarction is "inferior". This is the most pronounced effect.
There is no real significant difference between the other sites.
Patients who were on prior beta blocker medication also have a lower
probability of survival.

